Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-06-28 Paper/paper.tex
s4 done

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
% \usepackage[switch,columnwise]{lineno} 
\usepackage{lineno}
\newcommand{\roughly}{\mathchar"5218\relax} % Different from \sim in spacing

@@

In the previous section, we showed that the effects of NS's deformation due to
its companion hole's tidal field become discernible in the GW spectrum only when
the following hold: (a) the BH has a sufficiently small mass, (b) BH spin
is positive aligned, i.e. $\chibh\gtrsim +0.4$, (c) the NS is not very compact,
with $\lambdans\gtrsim 1000$, and finally, (d) the source is close enough
such that SNR $\gtrsim 30$.
% 
The first two conditions both enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BHBH signals are maximal.
% 
The first three also reduce the onset frequency of NS's disruption, allowing
for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which would maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. We, therefore,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
turn the question around and ask: under similarly favorable circumstances, can
we gain insights about the internal structure of neutron stars from GW observations.

%
%Ultimately, the extent of tidal moduilations of emitted GWs depends on the
%equation of state of neutron star matter, which becomes an indirect
%measurable in GW astrophysics.
% 
%We therefore turn to the question: If we do observe GWs from disruptive 
%coalescing NSBHs, what information can we extract from them about the internal
%structure of neutron stars?
% 
In order to explore this, we calculate here the accuracy with which we measure
$\lambdans$ from {\it single} GW observations. We sample the same set of
disruptive NSBH systems as in the previous section, i.e. those with
$q=\{2,3,4,5\}$, $\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$.
For each unique combination of these parameters, we generate separate injections
at SNRs $\rho=\{20,30,50,70\}$, using the LEA+ model~\cite{Lackey:2013axa}. We
perform full Bayesian analyses on each of these injections using LEA+ templates,
and obtain probabilistic estimates for source binary parameters using GW information
alone (including the NS tidal deformability parameter $\lambdans$). Our priors
on component masses and spins remain as in the previous section, while $\lambdans$
is sampled uniformly from $[0,4000]$.

As an illustration, we show the recovered
probability distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix $q = m_1/m_2 =
5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, while varying $\lambdans$
over $\{1000, 1500, 2000\}$ between the three panels. The SNR is fixed at $\rho=50$.
The true value of $\lambdans$ is given the green (dashed) vertical line,
while the median recovered value is marked by a red (dashed) line. The darker
shaded region marks the $90\%$ confidence interval on $\lambdans$. One thing
to note in this figure is that the measurement uncertainty on $\lambdans$, i.e.
the full width of its $90\%$ confidence interval, grows slower than $\lambdans$
as $\lambdans$ is turned up (while the fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases with increasing $\lambdans$).
%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
$\rho=30$ (instead of $50$). In the top panel, we show the correlation between
non-tidal parameters, as manifested in their recovered joint probability
distributions.
In the middle row, we show the same between $\lambdans$ and non-tidal 
parameters, i.e. chirp mass (left panel), mass ratio $\eta$ (middle panel), and
BH spin $\chibh$ (right). The same is shown in the row below, with the
difference that the SNR has been turned up to $50$ to bring out the intrinsic
ambiguity contours.
%
We find that the measurement of $\lambdans$ is only weakly degenerate with
other parameters, and may only marginally improve if we knew non-tidal parameters
to better accuracy. The predominant factor that would improve the measurement
accuracy for $\lambdans$ is the signal strength, SNR.

%\red{Present posterior distributions of selected configurations first..}

%\red{%
%\begin{enumerate}
% \item State the prior on $\lambdans$ because $\lambdans$ weakly deforms waveforms,
%I would expect that measurement uncertainty in $\lambdans$ is approx. independent
%of $\lambdans$, except for running into the $\lambdans\geq 0$ prior.
% \item Is $\lambdans$ measurement degenerate with other parameters? Either cite 
% known studies [if any], or provide 2-d posterior plots.
%\end{enumerate}
%}

%The precision with which $\lambdans$ can be measured is given by the full 
%width of its $90\%$ confidence interval, as a fraction of the true value. 
In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the same as a function
of binary parameters and signal strength. Each panel shows the percent statistical
uncertainty $\dlambda$ in the measurement of $\lambdans$ as a function of BH 
masses and spins, with NS parameters and signal strength being held fixed. 
In each row, the signal strength is increased from
left to right with $\rho = 20\rightarrow 70$. In each column, the deformability
of the NS is increased from top to bottom, with $\lambdans=500\rightarrow 2000$.
% 
We find that at SNRs around $\rho=30$, we begin to constrain $\lambdans$ 
meaningfully with $\dlambda\approx 100-150\%$. Amongst other parameters, 
the BH spin and mass play a big role. A smaller BH with a larger spin always
allows for a more precise measurement on $\lambdans$. Visually we can see that
in each panel the bottom right corner, which corresponds to low-mass BHs
with large spins, is the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without getting remarkably lucky and
finding such a binary with $\rho\gtrsim 50$.

In Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we take the case of
moderately deformable neutron stars, with actual $\lambdans = 1000$ (left 
panel) and $1500$ (right panel), and ask how loud such a signal will have
to be for us to put meaningful constraints on the tidal deformability of NSs.
In the left panel, as a function BH mass and spin, 
we show the lowest SNR value $\rho_c$ in order for the signal to have enough
information that we can constrain $\dlambda$ below $200\%$.
%\red{%
%\begin{enumerate}
% \item Inconsistent notation $\delta$ vs $\dlambda$. Unify.
% \item Explain precise definition of $\delta\lambdans/\dlambda$: Is this the 
% full width or half width of the posterior. Such explanation fit in nicely into
% the discussion of the new $\lambdans$ posterior plots like Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}.
%\end{enumerate}
%}
%\hspace{-3mm}
Note that being able to constrain $\lambdans$ within a factor of $2$ interval 
(i.e. within $\pm 100\%$) is a fairly modest requirement. We find promising 
results, which suggest that there is a fairly wide range of BH spins, less
wide (but non-trivial) range of BH masses, within
which if we get a moderately loud NSBH signal with $\rho\approx 23$, we constrain
$\lambdans$ within a factor of $2$ interval (i.e. within $\pm 100\%$) around
the true value. While this is not very precise, it can possibly allow us to
rule out candidates for the equation of state of neutron star matter.
% 
If we want to constrain $\lambdans$ better with a single observation, however,
we need SNRs of $30$ or higher. This can be seen from the right
panel of Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0}, which is similar
to the left but for a higher $\lambdans (=1500)$ system, with a more stringent
constraint requirement that $\lambdans$ be measured within a factor of $1$ of 
its true value (i.e. with $\pm 50\%$ errors, instead of $\pm 100\%$).
\red{Dont use two different thresholds in figure.} Even for
the most conducive cases, we {\it need} the SNR to be above $30$ to reach our 
accuracy threshold (which is more than $3\times$ less likely to happen than
getting a signal with $\rho=20$).

%\red{%
%(i) Rework text, so that you slowly build up the information needed. First, explain
%the general idea pursued, then discuss one panel first, then introduce second
%panel, finally, contrast both panels.
%(ii) High information density. Distinction between left and right panels notation
%not needed in the first sentence. deferring this info by a few sentences will make
%it easier for the reader to follow the discussion.
%}
%\prayush{%
%Lastly, in Fig.~\ref{fig:TT_LambdaErrorCurves_BHspin_BHmass_CI90_0} we show the
%conditions that BH parameters must fulfill beside having an SNR of $20$ (left 
%panel) or $30$ (right panel), to obtain meaningful constraints on $\lambdans$.
%At $\rho=20$, the left panel shows that we need high BH spins 
%$\chibh\gtrsim +0.6$ and a very deformable NS with $\lambdans\geq 1500$, in 
%order to put a factor of $2$ bound on $\lambdans$. At $\rho=30$, with
%$\lambdans\gtrsim 1000$ and $\chibh\geq+0.6$ we can put a similar factor of
%$2$ bound on $\lambdans$, but with $\chibh\sim+0.75$ and an extremely deformable
%NS with $\lambdans=2000$ one could even put a $\pm 50\%$ error-bar on 
%$\lambdans$ measurement.
%}

In summary, with a single moderately loud BHNS observation, we could constrain
the NS compactness parameter $\lambdans$ within $\pm 100\%$ of its true value.
To measure better with one observation, we will need to get lucky with signals
with $\rho\geq 30$ and high BH spins, or $\rho\geq 50$.

>>>>

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
% \usepackage[switch,columnwise]{lineno} 
\usepackage{lineno}
\interfootnotelinepenalty=10000

\newcommand{\roughly}{\mathchar"5218\relax} % Different from \sim in spacing

@@

In the previous section, we showed that the effect of NS's deformation due to
its companion BH's quadrupolar tidal field becomes discernible in the
GW spectrum only when the following hold:
(a) the BH has a sufficiently small mass, (b) BH spin
is positive aligned, i.e. $\chibh\gtrsim +0.4$, (c) the NS is not very compact,
with $\lambdans\gtrsim 1000$, {\it and} (d) the source is close enough
such that its GW SNR $\gtrsim 30$.
% 
The first two conditions both enhance the tidal distortion of the star and increase
the number of orbits the system goes through at small separation, where the
differences between NSBH and BHBH signals are maximal.
% 
The first three also reduce the onset frequency of NS's disruption, allowing
for it to happen earlier in the orbit. 
% 
We expect that these conditions are also the ones which would maximize the
likelihood of {\it measuring} tidal effects in NSBH signals. In this section,
%
%In addition, a SNR of $30$ corresponds to a $1.35+5M_\odot$ binary with 
%$\chibh = 0.4$ at a luminosity distance of $XXX$~Mpc, if optimally aligned
%with LIGO ($XXX$~Mpc if the binary subtends a $45^o$ angle with LIGO's line of
%sight.
%
%These conditions are strong, but not unrealistic. For e.g., a detection rate
%of $20$ NSBHs a year implies $1-2$ of them will have SNRs$\gtrsim 30$, since
%the number of expected events scales as $1/\rho^2$ if we assume that sources
%are distributed uniformly in spacial volume.
%
we turn the question around and ask: under similarly favorable circumstances, can
we gain insights about the internal structure of neutron stars from GW observations.

%
%Ultimately, the extent of tidal moduilations of emitted GWs depends on the
%equation of state of neutron star matter, which becomes an indirect
%measurable in GW astrophysics.
% 
%We therefore turn to the question: If we do observe GWs from disruptive 
%coalescing NSBHs, what information can we extract from them about the internal
%structure of neutron stars?
% 
We begin by calculating the accuracy with which we measure
$\lambdans$ from {\it single} GW observations. We sample the same set of
disruptive NSBH systems as in the previous section, i.e. those with
$q=\{2,3,4,5\}$, $\chibh=\{-0.5,0,+0.5,+0.75\}$, and $\lambdans=\{500, 800, 1000, 1500, 2000\}$;
fixing the NS mass $\mns=1.35 M_\odot$ and $\chins=0$.
For each unique combination of these parameters, we generate separate 
zero-noise injections at SNRs $\rho=\{20,30,50,70\}$, using the LEA+ model. We
perform full Bayesian analyses on each of these injections using LEA+ templates,
and obtain probabilistic estimates for source binary parameters using GW information
alone (including the NS tidal deformability parameter $\lambdans$). Our priors
on component masses and spins remain as in the previous section, while $\lambdans$
is sampled uniformly from $[0,4000]$.

As an illustration, we show the recovered
probability distribution for $\lambdans$ for three specific configurations in 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. We fix $q = m_1/m_2 =
5.4M_\odot/1.35M_\odot = 4$, with $\chibh=+0.5$, while varying $\lambdans$
over $\{1000, 1500, 2000\}$ between the three panels. The SNR is fixed at $\rho=50$.
The true value of $\lambdans$ is given the green (dashed) vertical line,
while the median recovered value is marked by a red (dashed) line. The darker
shaded region marks the $90\%$ confidence interval on $\lambdans$. One thing
to note in this figure is that the measurement uncertainty on $\lambdans$, i.e.
the full width of its $90\%$ confidence interval, grows slower than $\lambdans$
as $\lambdans$ is turned up (while the fractional uncertainty
\begin{equation}
\delta\lambdans^{90\%}:= (\Delta\lambdans)^{90\%}/\lambdans
\end{equation}
decreases with increasing $\lambdans$).
%
Next, in Fig.~\ref{fig:SingleSystemLambda2DPDFs} we show the correlation of
mass, spin, and tidal parameter measurements. We keep the binary parameters
as in Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}, with $\lambdans=2000$, and
$\rho=30$ (instead of $50$). In the top panel, we show the correlation between
non-tidal parameters, as manifested in their recovered joint probability
distributions.
In the middle row, we show the same between $\lambdans$ and non-tidal 
parameters, i.e. chirp mass (left panel), mass ratio $\eta$ (middle panel), and
BH spin $\chibh$ (right panel). The same is shown in the row below, with the
difference that the SNR has been turned up to $\rho=50$, which brings out the
intrinsic ambiguity contours more clearly.
%
We find that the measurement of $\lambdans$ is only weakly degenerate with
other parameters, and may only marginally improve if we knew non-tidal parameters
to better accuracy. The predominant factor that would improve the measurement
accuracy for $\lambdans$ is the signal strength, SNR.

In Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR} we show the main results
of this section. In each panel, as a function of black hole mass and spin,
we show the measured $90\%$ confidence intervals for $\lambdans$, i.e.
$(\Delta\lambdans)^{90\%}$. These correspond to the full width of the dark
shaded region in the illustrative figure,
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR}. The effect of signal strength,
i.e. SNR, can be seen as we go from left to right in each row. The
effect of NS's tidal deformability $\lambdans$ on its own measurability
can be seen by comparing panels within each column. The NS becomes more 
deformable as we go from top to bottom. In addition, contours of
$\dlambda=\{50\%,75\%,100\%,150\%,200\%\}$ are also shown in all panels.
%
A clear pattern emerges in the left-most column, which corresponds to
$\rho=20$. We find that at this signal strength, our measurement of 
$\lambdans$ is dominated by the width of our prior on $\lambdans$. In other
words, our measured confidence interval is restricted by the
allowed range of $\lambdans$. Increasing the signal strength to
$\rho=30$ gives somewhat similar results, as can be seen from the second-from-left
column of the figure. We do, however, find more encouraging results
for high BH spins, in which case we can put $\pm 100\%$ to $\pm 75\%$
bounds on $\lambdans$~\footnote{The symmetric error-bars of
$\pm\mathrm{X}\%$ correspond to $\dlambda = 2\mathrm{X}\%$.}.
In addition to large BH spins, this level of
precision will also require rather deformable neutron stars.
%
At higher SNRs, $\rho\gtrsim 50$, we can put stronger constraints
on $\lambdans$. With a {\it single} observation of a $q=4$ binary with
$\chibh\geq 0.6$ and $\rho = 50$~\footnote{For an optimally oriented
source with $q=4, \mns=1.35M_\odot, \chibh=0.6$, an SNR of 
$\rho = 50$ corresponds to a luminosity distance of $\approx 113$Mpc.},
we would be able to estimate 
$\lambdans$ to within $\pm 40\%$ of its true value 
(which is equivalent to measuring the ratio of its radius to mass
with an uncertainty of about $\pm 10\%$).

Amongst other source parameters, BH mass and spin play a dominant role. A smaller
BH with a larger spin always allows for a more precise measurement on $\lambdans$.
Visually we can see this in the bottom right corner of each panel in
Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}, which corresponds to low-mass BHs
with large spins, and is simultaneously the region of smallest measurement errors on $\lambdans$.
The actual deformability of the NS also plays an important role on its own
measurability. For e.g., when $\lambdans\leq 1000$, it is fairly difficult
to meaningfully constrain $\lambdans$ without requiring the source to be
close ($\approx 100$Mpc) with a GW SNR $\rho\gtrsim 50$. Quantifying this further,
in Fig.~\ref{fig:TT_SNRThresholds_BHspin_BHmass_CI90_0} we show the minimum
signal strength required to attain a certain level of confidence in our
$\lambdans$ measurement, as a function of BH properties. The NS is allowed
the most favorable (hardest) EoS considered, with $\lambdans^\mathrm{true}=2000$.
% 
We first note that, even with the most favorable BH and NS properties, achieving
a $\pm 50\%$ measurement certainty on $\lambdans$ will require a GW SNR
$\rho\gtrsim 30$. If we additionally restrict BH masses to lie outside of the so-called
astrophysical mass-gap~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,
Littenberg:2015tpa}, we will simultaneously need to restrict BH spins
to $\chibh\gtrsim +0.5$ to obtain the same measurement confidence at the same
source location.

In summary, with a single moderately loud ($\rho\lesssim 30$) GW signal from
a disruptive BHNS coalescence, we can constrain
the NS compactness parameter $\lambdans$ within $\pm 100\%$ of its true value.
To measure better with one observation, we will need a more fine-tuned source, with
$\rho\geq 30$ and high BH spins, or $\rho\geq 50$.
% 
Finally, we note that these results are {\it conservative}, and 
BHs with spins $\chibh > 0.75$ will prove to be even more favorable laboratories
for $\lambdans$ measurement. However, we are presently unable to explore this case
in quantitative detail due to waveform model restrictions~\cite{Lackey:2013axa},
which will restrict our analyses of actual GW signals as well.
